Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-08 paper/paper.tex
Finished the section on basic path indexes.

\subsection{Graphs}

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for $0 \le i < \abs{P}-1$. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$.

We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$ not used elsewhere in the graph. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$.

% FIXME example of strings, DBG, long paths, false positives
Let $\mathcal{S}$ be a (potentially infinite) collection of strings over alphabet $\Sigma$, each of them ending with $k$ copies of the endmarker $\$$. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$, where each distinct \emph{\kmer{k}} (substring of length $k$) $X$ occurring in collection $\mathcal{S}$ is represented by a node $v_{X} \in V$. Each node $v_{X}$ has a \emph{key} $G.\gkey(v_{X}) = X$ and a label $G.\glabel(v_{X}) = X[0]$. The edges of the de~Bruijn graph represent \kmer{k+1}s $Y$ occurring in the collection. For each such substring $Y = cX$, where $c \in \Sigma$, we have an edge from node $v_{Y[0, k-1]}$ to node $v_{X}$. We use the node corresponding to substring $\$^{k}$ as the sink node, and add the source node (with $G.\gkey(s) = \#$) and the associated edges in the usual way. We may also have edges from the source node to the nodes representing the prefixes of the collection.

\begin{lemma}[No false negatives]
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ not containing character $\$$. Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
\end{lemma}

@@

\begin{lemma}[No false positives with $\abs{P} \le k+1$]
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$. Then path $P$ either contains the source node $s$ or its label is a substring of a string $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
Assume that path $P = v_{0} \dotsm v_{\abs{P}-1}$ does not contain node $s$. If $\abs{P} = 0$, the label is an empty string and hence a substring of each string $S \in \mathcal{S}$. If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$.

Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma \setminus \set{\#}$ and $1 \le \abs{X} \le k$. By induction, string $X$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$. Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
\end{proof}

@@

% FIXME LF(i, c) picture from the slides for the earlier DBG example
The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and \emph{de Bruijn graphs}. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the paths starting from them. The transformation is expensive and may increase the size of the graph exponentially. After sorting the nodes of the transformed graph, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes into BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges into the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

The design of a path index is a trade-off between the maximum supported query length and the size of the index. In the worst case, a graph may have up to $(\sigma-2)^{k}$ distinct labels for paths of length $k$ (ignoring paths containing the source node or the sink node), and the length of possible paths is unlimited in cyclic graphs. GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary. The price of this was the possibility of \emph{false negatives}: path labels that exist in the graph but not in the index.

% FIXME from here
k-mer indexes based on hash tables, sorted k-mers

similarities between k-mer indexes and de Bruijn graphs: order-k+1 index can simulate order-k DBG; order-k DBG can be used as an order-k index

define de Bruijn graphs for graphs (infinite strings, no source node, repeat sink node instead)

using de Bruijn graphs as indexes; false positives
- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
- add values (pointers to start nodes) to each node
- valid for existence queries of length up to k+1 and locate queries of length up to k

>>>>

\subsection{Graphs}\label{sect:graphs}

@@

The graphs we consider are \emph{labeled} with alphabet $\Sigma$: each node $v \in V$ has a \emph{label} $G.\glabel(v) \in \Sigma$. A \emph{path} in a graph is a sequence of nodes $P = v_{0} \dotsm v_{\abs{P}-1}$ such that $(v_{i}, v_{i+1}) \in E$ for all $i, i+1 \in \set{0, \dotsc, \abs{P}-1}$. The label of a path is the concatenation of node labels $G.\glabel(P) = G.\glabel(v_{0}) \dotsm G.\glabel(v_{\abs{P}-1})$. We can also generalize the definition for infinite paths $P = (v_{i})_{i \in Z}$, where set $Z$ is one of $\mathbb{Z}$, $\set{x \in \mathbb{Z} \mid x \ge 0}$, and $\set{x \in \mathbb{Z} \mid x \le 0}$. The labels of such paths are infinite sequences of characters instead of strings.

We assume for convenience that all graphs have two special nodes: the \emph{source} node $s$ and the \emph{sink} node $t$. To distinguish them from the other nodes, we label them with characters $G.\glabel(s) = \#$ and $G.\glabel(t) = \$$. The label of the sink node is unique in the graph, while the label of the source node can be used in other nodes for technical purposes. We add an edge $(s, v)$ for all nodes $v \in V \setminus \set{s}$ that have no other incoming edges, and an edge $(v, t)$ for all nodes $v \in V \setminus \set{t}$ with no other outgoing edges. We also add the edge $(t, s)$ to guarantee that $G.\gindegree(v) \ge 1$ and $G.\goutdegree(v) \ge 1$ for all nodes $v \in V$. However, we assume for convenience that no path in the graph crosses the edge $(t, s)$.

% FIXME example of strings, DBG, long paths, false positives
Let $\mathcal{S}$ be a collection of strings over alphabet $\Sigma$. We assume that each string $S \in \mathcal{S}$ begins with $k$ copies of $\#$ and ends with $k$ copies of $\$$, with no other occurrences of these characters in the string. The \orderk{k} \emph{de~Bruijn graph} of $\mathcal{S}$ is a graph $G = (V, E)$, where each distinct \emph{\kmer{k}} (substring of length $k$) $X$ occurring in collection $\mathcal{S}$ is represented by a node $v_{X} \in V$. Each node $v_{X}$ has a \emph{key} $G.\gkey(v_{X}) = X$ and a label $G.\glabel(v_{X}) = X[0]$. The edges of the de~Bruijn graph represent \kmer{k+1}s $Y$ occurring in the collection. For each such substring $Y = cX$, where $c \in \Sigma$, we have an edge from node $v_{Y[0, k-1]}$ to node $v_{X}$. We use the node corresponding to substring $\#^{k}$ as the source node $s$ and the node corresponding to $\$^{k}$ as the sink node $t$, adding the technical edge $(t, s)$ in the usual way.

\begin{lemma}[No false negatives]\label{lemma:dbg-fn}
Let $G = (V, E)$ be the \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $X$ be a substring of $\mathcal{S}$ containing at most one $\$$. Then there is a path $P$ in the graph with $G.\glabel(P) = X$.
\end{lemma}

@@

\begin{lemma}[No false positives with $\abs{P} \le k+1$]\label{lemma:dbg-fp}
Let $G = (V, E)$ be an \orderk{k} de~Bruijn graph of string collection $\mathcal{S}$, and let $P$ be a path in $G$ with $\abs{P} \le k+1$. Then the label of the path is a substring of a string $S \in \mathcal{S}$.
\end{lemma}

\begin{proof}
If $\abs{P} = 0$, the label is an empty string and hence a substring of each string $S \in \mathcal{S}$. If $\abs{P} = 1$, the label is equal to $G.\glabel(v_{0})$, which is a prefix of $G.\gkey(v_{0})$ and hence a substring of $\mathcal{S}$. Now assume that $G.\glabel(P) = cX$, where $c \in \Sigma$ and $1 \le \abs{X} \le k$. By induction, string $X$ is a prefix of $G.\gkey(v_{1})$. As edge $(v_{0}, v_{1})$ exists, $c \cdot G.\gkey(v_{1})$ must be a substring of $\mathcal{S}$. Hence $cX$ is a substring of $\mathcal{S}$, and also a prefix of $G.\gkey(v_{0})$ if $\abs{cX} \le k$.
\end{proof}

@@

% FIXME LF(i, c) picture from the slides for the earlier DBG example
The \emph{generalized compressed suffix array} (GCSA) \cite{Siren2014} is a further generalization of the XBW transform for a class of graphs that includes \emph{directed acyclic graphs} and de Bruijn graphs. Before a graph can be indexed, we have to transform it into an equivalent graph, where the nodes can be unambiguously sorted by the labels of the paths starting from them. The transformation is expensive and may increase the size of the graph exponentially. After sorting the nodes of the transformed graph, we encode them using three sequences. $\BWT$ contains the labels of the predecessors, while bitvectors $\bvIN$ and $\bvOUT$ encode the indegrees and outdegrees of each node in the same way as bitvector $B$ of the XBW transform. \LFmapping{} uses select queries on bitvector $\bvIN$ to map nodes into BWT ranges, ordinary \LFmapping{} with $\BWT$ to map incoming edges into the corresponding outgoing edges, and rank queries on bitvector $\bvOUT$ to map the outgoing edges to the predecessor nodes.

@@

The design of a path index is a trade-off between the maximum supported query length and the size of the index. In the worst case, a graph may have up to $\sigma^{k}$ distinct labels for paths of length $k$ (ignoring the special treatment of characters $\#$ and $\$$), and the length of possible paths is unlimited in cyclic graphs. GCSA \cite{Siren2014} tried to avoid this by considering only a relatively simple class of graphs, using \emph{pruning heuristics} to simplify the graphs when necessary. The price of this was the possibility of \emph{false negatives}: path labels that exist in the graph but not in the index.

\subsection{Basic indexes}

The \emph{\kmer{k} index} is the simplest path index. In its most basic form, the \kmer{k} index consists of a parameter value $k$, a hash function $h$, and an array $A$ of \emph{key-value pairs} $(X, W)$, where $X \in \Sigma^{k}$ and $W \subseteq V$. Given a \kmer{k} $X \in \Sigma^{k}$, we start looking for key $X$ from $A[h(X)]$, and list the set of start nodes $W$ as matches if we find it. Searching using a hash table is fast, but we can only search for patterns of length $k$, and the array requires a lot of space.

Another representation of the \kmer{k} index trades query performance for the ability to search for shorter patterns. Instead of using a hash table, we sort the key-value pairs by their keys, and store the pairs in array $A$ in sorted order. Given a pattern $X \in \Sigma^{\ast}$, $\abs{X} \le k$, we use binary search to find the range $A[sp, ep]$ of pairs $(X_{i}, W_{i})$, where pattern $X$ is a prefix of the key $X_{i}$. We then list the union $\bigcup_{i=sp}^{ep} W_{i}$ as the set of matches.

As de~Bruijn graphs and \kmer{k} indexes are both based on \kmer{k}s, we can use one to simulate the other. Before proceeding, we first have to define the de~Bruijn graph of a labeled graph. Let $P$ be a (finite or infinite) path in graph $G = (V, E)$. We say that path $P$ is \emph{maximal}, if starts at the source node or extends infinitely to the left, and ends at the sink node or extends infinitely to the right. We construct a collection of character sequences $\mathcal{S}$ by taking all maximal paths $P$ in graph $G$ and creating a sequence $S$ based on $G.\glabel(P)$ for each such path. If path $P$ is infinite in both directions, we use the label directly. If the path starts at the source node, we adjust the number of characters $\#$ at the beginning to match the order $k$ of the de~Bruijn graph. Similarly, if the path ends at the sink node, we adjust the number of characters $\$$ at the end. We can then generalize the definition of de~Bruijn graphs in Section~\ref{sect:graphs} to create a de~Bruijn graph of sequence collection $\mathcal{S}$.

Given a \kmer{k+1} index of graph $G = (V, E)$, we can use it to simulate the \orderk{k} de~Bruijn graph $G' = (V', E')$ of $G$. As the nodes of the de~Bruijn graph correspond to \kmer{k}s and the edges correspond to \kmer{k+1}s, we can represent the nodes by the adjacent edges. If we are interested in node $v_{X} \in V'$, we search for incoming edges $cX$ and outgoing edges $Xc$ for all $c \in \Sigma$. Determining the existence of a node requires $2 \sigma$ queries in the \kmer{k+1} index, and we learn the presence of adjacent nodes and edges in the process.

On the other hand, we can use the \orderk{k} de~Bruijn graph $G' = (V', E')$ of graph $G = (V, E)$ directly as a \kmer{k} index of $G$. For each key-value pair $(X, W)$ in the index, we find the node $v_{X} \in V'$ with $G'.\gkey(v_{X}) = X$, and attach the \emph{value} $G'.\gvalue(v_{X}) = W$ to the node. We can then search for patterns of any length in the de~Bruijn graph. As Lemmas~\ref{lemma:dbg-fn} and \ref{lemma:dbg-fp} suggest, there will be no false negatives, and no \emph{false positives} with short enough patterns. With longer patterns, we have to \emph{verify} the results in graph $G$ to avoid false positives. We will develop the idea of using graphs similar to de~Bruijn graphs as path indexes further in the next section.

\subsection{Path graphs}

% FIXME from here
using de Bruijn graphs as indexes; false positives
- string Y matches node with key X: X is a prefix of Y or Y is a prefix of X
- valid for existence queries of length up to k+1 and locate queries of length up to k
